Finally Tagless, Partially Evaluated

### 2007-06-12 tagless/tagless-final.tex
shorten, shorten, shorten

Our main idea is to encode
de Bruijn or higher-order abstract syntax
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

\begin{quote}
\small
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{quote}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser\slash type\hyp checker reading from a text string.
We leave aside the solved problem of writing such a parser\slash type\hyp checker,
whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.

@@

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let test1 = A (L (V VZ), B true)}.
%aplas-else
\begin{comment}
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}
\end{comment}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions
\begin{code3}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code3}
which comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let testf1 = app (lam varZ) (b true)}
%aplas-else
\begin{comment}
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\end{comment}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, |testf1 ()|, obtaining |true|.
%aplas-else
\begin{comment}
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\end{comment}
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
%aplas-specific
open term such as \texttt{app (lam (varS varZ)) (b true)}
gives a type- rather than a run-time error.
%aplas-else
\begin{comment}
\begin{code}
open term such as |testf3| below gives a type error rather than
a run-time error
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\end{comment}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

@@

Our solution is emphatically distinct from Church-encoding of the
universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
requires two continuations; the interpreter like |app| above would
have to provide both to the encoding of |e1|. The continuation
corresponding to the |UB| branch of |u| must either raise an error or
loop. For a well-typed object term that error continuation is never
invoked, yet it must be supplied. Our interpreter has no error
continuations at all.

The evaluator above is wired directly into the
functions |b|, |lam|, |app|, and so on.  In the rest of this paper, 
we explain how to abstract the interpreter so as
to process the same term in many other
ways: compilation, partial evaluation, size
computation, and so forth.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to the most severe lack of space, we cannot describe
these features and refer the reader to the accompanying code.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnotemark[\value{footnote}]
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to express Fibonacci, factorial, and
power.

In contrast to \S\ref{intro}, we encode binding using higher-order
abstract syntax (HOAS) \cite{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, 
such as if we forget |app| somewhere above.

\begin{figure*}[t]
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c, 'dv) repr

  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

module EX(S: Symantics) = struct
  open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
                         if_ (leq n (int 0)) (fun () -> int 1)
                             (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |c|, an
environment classifier \cite{WalidPOPL03} (required by MetaOCaml for
code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
%aplas: inlining
\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
\begin{comment}
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
\end{comment}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code3}
repr_pe Int      = Int          repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code3}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int : int  -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation (\texttt{bool}, \texttt{mul},
\texttt{leq} are elided)}
\label{fig:ocaml}
\end{figure*}

@@

  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} are elided)}
\label{fig:pe}
\end{figure}

>>>>

Our main idea is to encode
% de Bruijn or higher-order abstract syntax
HOAS
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

\begin{quote}
\small
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm John Reynolds~\citep{reynolds-definitional}
\end{quote}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
%aplas
\begin{comment}
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser\slash type\hyp checker reading from a text string.
We leave aside the solved problem of writing such a parser\slash type\hyp checker,
whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
\end{comment}
%aplas-else
We leave aside the solved problem of writing a parser\slash type\hyp checker,
for embedding object language objects into the metalanguage
(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
and just enter them by hand.
%aplas

@@

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml~%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let test1 = A (L (V VZ), B true)}.
%aplas-else
\begin{comment}
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}
\end{comment}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions
\begin{code4}
let varZ env    = fst env       let b (bv:bool) env = bv
let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
let lam e env   = fun x -> e (x,env)
\end{code4}
which comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let testf1 = app (lam varZ) (b true)}
%aplas-else
\begin{comment}
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\end{comment}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, |testf1 ()|, obtaining |true|.
%aplas-else
\begin{comment}
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\end{comment}
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
%aplas-specific
open term such as \texttt{app (lam (varS varZ)) (b true)}
gives a type- rather than a run-time error.
%aplas-else
\begin{comment}
\begin{code}
open term such as |testf3| below gives a type error rather than
a run-time error
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\end{comment}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

@@

Our solution is emphatically distinct from Church-encoding of the
universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
requires two continuations; the interpreter (see |app| above) would
have to provide both to the encoding of |e1|. The continuation
corresponding to the |UB| branch of |u| must either raise an error or
loop. For a well-typed object term that error continuation is never
invoked, yet it must be supplied. Our interpreter has no error
continuation at all.

The evaluator above is wired directly into the
functions |b|, |lam|, |app|, and so on.  %In the rest of this paper, 
We explain how to abstract the interpreter so as
to process the \emph{same} term in many other
ways: compilation, partial evaluation, CPS
conversion, and so forth.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnote{again, see code}
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to express Fibonacci, factorial, and
power.
In contrast to \S\ref{intro}, we encode binding using higher-order
abstract syntax (HOAS) \cite{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

@@

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, 
such as if we forget |app| somewhere above.

\begin{figure*}[t]
\begin{floatrule}
\begin{code}
module type Symantics = sig type ('c, 'dv) repr
  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

module EX(S: Symantics) = struct open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
                         if_ (leq n (int 0)) (fun () -> int 1)
                             (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}
To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |c|, an
environment classifier \cite{WalidPOPL03} (required by MetaOCaml for
code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
%aplas: inlining
\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
\begin{comment}
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
\end{comment}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by\vspace*{-0.8em}
\begin{code3}
repr_pe Int      = Int          repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code3}
\vspace*{-0.8em}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int  -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation (\texttt{bool}, \texttt{mul},
\texttt{leq} are elided)}
\label{fig:ocaml}
\end{figure*}

@@

  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
                                (fun () -> abstr (ee ())))
  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} are elided)}
\label{fig:pe}
\end{figure}
